Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-07-16 Manuscript/manuscript.tex
Reorganization of discussion
Some reorganization of discussion to put paragraphs on the roles of low
and high quality patches first.  Some new transitions and notes on
additional discussion needed.

\textcolor{red}{Need to emphasize novelty of this work, especially compared to Hess, Gog, and Park.  Discuss that movement in this model is not only heterogeneous, but also directional and what that means.  Also consider reorganizing to put the more interesting elements at the front, and get rid of the less interesting parts (i.e. the first paragraph).  As with notes for results and methods, need to be clear about which analyses support which claims, and specifically need to be careful about discussing the different patch quality distributions and distinguishing between the symmetric and asymmetric distributions.}

The distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.  These impacts are driven largely by the contrasting roles played by low and high quality habitat and their relative balance in the metapopulation.

This work, together with previous studies of disease dynamics in a metapopulation (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}), suggests that there is a general trade-off between forces that facilitate metapopulation stability at the expense of disease spread, and forces that hinder disease spread at the expense of metapopulation stability.  The position of these trade-offs, and their relative importance depend largely on the strength and structure of pathogen transmission.

@@

Though beyond the scope of the current study, future work could more deeply investigate the effect of changing movement rates across different patch quality distributions.

By modeling the spatial spread of an environmentally-transmitted pathogen, these simulations suggest that highly heterogeneous metapopulations are more prone to widespread pathogen spread than more homogeneous metapopulations (Fig~\ref{poutcome}).  In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).  The mechanisms underlying this behavior can be explored by drawing parallels with models of pathogen spread through a contact network.  In this context, our metapopulation can be viewed as a network, with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).

However, by including environmental transmission, our models suggest that the overall effect of patch quality variance depends on the structure of the metapopulation, the resulting colonization and extinction rates, and the relative balance of the effects of low and high quality patches.  In the baseline model, the increase in pathogen spread in high variance metapopulations results primarily from the presence of high quality patches and their ability to overwhelm low quality patches at high longevity through a combination of their relative stability, the high connectivity due to the uniform spatial structure, and the relative strength of direct infection.  The assumptions tested in the other models all weaken these effects -- by constraining movement through the lattice structure; by limiting the influence of direct infection and thus the influence of colonists from infected high quality patches; and by reducing the stability of high quality patches relative to low -- thus preventing low quality patches from being overwhelmed by the high quality patches.

The differences observed between the different parameterizations of the model highlight the fact that the effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches on pathogen spread and metapopulation stability.  Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.  These two properties have important implications for pathogen spread in the metapopulation. Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure on high quality patches results from both transmission from the environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

\textcolor{red}{High qualily patches also experience more infection events because they are able to support susceptibles long enough for them to get infected, whereas susceptibles on low quality patches are probably going to go extinct before they can become infected (though the effect persists when $\alpha = 0$ and is substantially diminished in the lattice model and when $\delta = 0.1$, so it seems that it is driven largely by connectivity and direct infection}

This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  Once the pathogen infects high quality patches, the larger number of host colonists produced by these patches allow it to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infectious occupancy in the metapopulation (Fig~\ref{sens}).

And \cite{Hess1996} who finds that implementing a quarantine on centralized patches reduces pathogen spread.

In contrast, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).  \textcolor{red}{Maybe also note that natural mortality on low quality patches is high enough that there's not enough time for susceptibles to to get infected.}

Though this model provides evidence that high quality patches can serve as traps and superspreaders, and low quality patches can serve as susceptible refuges, the effects of these processes on total metapopulation occupancy depend on the strength and structure of the two transmission pathways.  \textcolor{red}{Need more here?}  In particular, these findings suggest the presence of a trade-off between facilitating metapopulation stability and inhibiting disease spread, with a pathogen's environmental longevity and ease of spread controlling the roles of different quality patches with respect to this trade-off.  When pathogen spread is limited or slow (e.g. due to low longevity, or weak direct transmission), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy.

\textcolor{red}{Maybe lay out the big picture in this paragraph: that there are essentially three different scenarios, very low pathogen spread, intermediate/endemic pathogen spread, and easy/pandemic pathogen spread and then go through each one and its consequences on the roles played by high and low quality habitat.}

Also discuss \cite{Sharp2011}, who show that management actions that increase carrying capacity can increase disease spread and destabilize population.

However, as pathogen longevity increases, the trade-off shifts and the addition of high quality patches is a net benefit to overall occupancy, reflecting the findings of \cite{Gog2002}, who find that under background infection from an alternative host, increasing movement rarely reduces occupancy, despite facilitating pathogen spread.  \cite{Park2012} further develops this work by noting that the magnitude of the detrimental effects of increased moevement on occupancy depends on the relative strength of different transmission pathways.  Specifically, he finds that as environmental transmission becomes stronger (e.g. higher longevity) relative to direct transmission, then total occupancy increases monotonically with movement, i.e. movement is a net benefit to the metapopulation.  These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. when the pathogen is long-lived in the environment), factors that facilitate metapopulation stability (i.e. increased movement, high quality patches) are a greater benefit than factors that inhibit pathogen spread (i.e. decreased movement, low quality patches).

>>>>

The distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.  These impacts are driven largely by the contrasting effects of low and high quality habitat on pathogen transmission, and their relative balance in the metapopulation. \textcolor{red}{Need more here? Possibly discussing some of the background colonization, extinction processes that are modified by patch quality, and how generally those interact with environmental transmission.}

Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.  These two properties have important implications for pathogen spread in the metapopulation. Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure on high quality patches results from both transmission from the environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  Once the pathogen infects high quality patches, the larger number of host colonists produced by these patches allow it to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infectious occupancy in the metapopulation (Fig~\ref{sens}).

The importance of high quality habitat in maintaining pathogen spread suggests that high quality habitat may be a good target for control efforts, e.g. quarantine or vaccination.  This echoes the insights of \cite{Hess1996} who simulated the spread of a directly transmitted pathogen through a metapopulation and found that implementing a quarantine on centralized patches reduced pathogen spread.  This insight is even more important for environmentally transmitted pathogens.

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).  \textcolor{red}{Maybe also note that natural mortality on low quality patches is high enough that there's not enough time for susceptibles to to get infected.}

Though this model provides evidence that high quality patches can serve as traps and superspreaders, and low quality patches can serve as susceptible refuges, the effects of these processes on total metapopulation occupancy depend on the strength and structure of the two transmission pathways.  \textcolor{red}{Need more here?}  In particular, these findings suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.  The position of these trade-offs, and their relative importance depend largely on the strength and structure of the different avenues of pathogen transmission.  Broadly, there are three different scenarios: very weak pathogen spread, intermediate/endemic pathogen spread, and easy/pandemic pathogen spread.  When pathogen spread is limited or slow (e.g. due to low longevity, or weak direct transmission), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.

Also discuss \cite{Sharp2011}, who show that management actions that increase carrying capacity can increase disease spread and destabilize population

As pathogen longevity increases, the trade-off shifts and the addition of high quality patches is a net benefit to overall occupancy, reflecting the findings of \cite{Gog2002}, who find that under background infection from an alternative host, increasing movement rarely reduces occupancy, despite facilitating pathogen spread.  \cite{Park2012} further develops this work by noting that the magnitude of the detrimental effects of increased moevement on occupancy depends on the relative strength of different transmission pathways.  Specifically, he finds that as environmental transmission becomes stronger (e.g. higher longevity) relative to direct transmission, then total occupancy increases monotonically with movement, i.e. movement is a net benefit to the metapopulation.  These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. when the pathogen is long-lived in the environment), factors that facilitate metapopulation stability (i.e. increased movement, high quality patches) are a greater benefit than factors that inhibit pathogen spread (i.e. decreased movement, low quality patches).

@@
